Model to study co-evolution of trade an culture

### 2015-04-02 article.tex
comments and fixes

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
\jm{to keep it simple, can we say ``evolutionary theory" instead of ``genetic theory" ?}
These analysis are often based on the genetic theory of neutral evolution \cite{neiman_stylistic_1995} and they take neutrality as a null-hypotheses, i.e. variations of traits are strictly explained by their frequency distribution in the population. There exist a large number of works using this model as the starting point to analyse patterns of cultural change in past societies \cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

@@

In particular, the trade of cultural products within a market system can potentially introduce a dynamics content-related bias linked to the price of the different products. This would also affect the frequency of the product within the population, which in turn would modify its price following a co-evolutionary dynamic.

@@

@@

However, a limitation of the previous works is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the data analysed. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data which is then qualitatively compared to the real historical data. By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data. The Agent Based Models (ABM) methodology offers a way to produce this kind of model~\cite{lake_trends_2014}.

@@

@@

@@

@@

@@

@@

@@

The first set of implementations of the core processes aims at reproducing a common result from the literature on cultural change.
In this article, the cultural change hypothesis tested is the neutral hypothesis, which has been shown to be followed by multiple cultural phenomenons~\cite{bentley_random_2004,bentley_specialisation_2005,mesoudi_random_2009}. Under this hypothesis, the production of goods is not tested, as a consequence, in this set of implementation, the \emph{production} process does not modify the content of the quantity vector of the agents.
The \emph{imitation} process used is  ``random copy'', which has been implemented as follow: each agent has a low probability to pick randomly one agent among all and copy its vector of values. The \emph{innovation} process, termed ``unbounded'', is triggered with a low probability and draw a new random value to replace an element $v^i_j$. The imitation and innovation probabilities are presented with other parameters in table~\ref{tab:parameters}.

The neutral hypothesis states that the ``random copy'' and the ``unbounded'' innovation process used under a fixed population size, leads to a repartition of cultural variants termed ``power law''. This distribution is characterized by a small number of very frequent traits and a large number of very rare traits. Since this hypothesis has been verified in previous works implementing similar \emph{imitation} and \emph{innovation} mechanisms, we expect in this work the observation of a similar ``power law' distribution. More formally, this distribution is formalised as : $$P(v)=C/v^\alpha $$, where $v$ is the number of time a variant has been repeated, $P(v)$ the probability to find that variant, $C$ a constant, $\alpha$ a variable.

@@

@@

The \emph{score} $s^i$ of an agent $i$ reflects the ability of this agent to obtain the products it needs. It is maximum when the quantity vector of an agent is equal to its need vector and lower proportionally to the distance between the need vector and the quantity vector.  It is formally computed as follow for agent $i$:

@@

This function ensures that each good count for the same proportion in the final score (i.e.: managing to get only the right amount of a good with a high ``need'' value will not give a better score to the agent). The values of $s^i$ obtained for different need values are depicted in figure~\ref{fig:fit}.

@@

@@

On top of producing goods, the agents have the ability to exchange them. During the trading phase the value associated to a good by an agent corresponds to the subjective price of the good for this agent. Briefly summarised, for each good that it does not produce, an agent will trade with the first parter that offer an acceptable trade, i.e. the agent that proposes a satisfiable ratio between the other product and the product produced by the agent.

@@

@@

\todo{upload and put an url, complete cite mare nostrum, put nb of hours per run (or the total)}
All experiments are performed with the simulator pandora~\cite{wittek_scalable_2012}. We have make the source code of our project available online~\footnote{put ulr}. All simulations presented are performed in the MareNostrumIII super computer~\cite{}, and took XXXh per run.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{ll|llll}
		\multicolumn{2}{r}{Agents Number}&\multicolumn{2}{c}{Our result}&\multicolumn{2}{c}{Bentley et al 2004}\\
			&$\mu$ & $\alpha$ & SD&$\alpha$&SD\\\hline
		N = 250	&0.004&1.53&0.03&1.54&0.02\\
			&0.008&1.55&0.03&1.55&0.01\\
			&0.016&1.57&0.02&1.57&0.01\\
			&0.064&1.66&0.01&1.67&0.01\\\hline
		N= 500	&0.004&1.50&0.02&1.53&0.03\\
			&0.008&1.52&0.02&1.57&0.01\\
			&0.016&1.55&0.03&1.61&0.04\\
			&0.064&1.78&0.08&1.81&0.10\\
	\end{tabular}
	\caption{Mean \& SD are calculate on 100 run.}
	\label{tab:mualpha}
\end{table}

@@

@@

@@

%The questions we would like to ultimately study are for example: how does the hierarchical structure changes the adoption of a new product ? How delays and interruptions in the trade network change the culture of a society ?

@@

>>>>

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
These analysis are often based on the theory of neutral evolution \cite{neiman_stylistic_1995}, which explains variation in the distribution of traits as a function of their frequency distribution in the population. Thus, a trait is assumed to lack any selective bias and its transmission will depend on its popularity amongst the rest of individuals. The comparison of a given frequency distribution found in data against this unbiased transmission model is traditionally used as a null-hypothesis testing for selection. There exist a large number of works using this approach as the starting point to analyse patterns of cultural change in past societies \cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

@@

In particular, trade within a market system can potentially introduce a complex dynamic bias to the neutral model. Trasmission will be linked to the price of the different products. This would also affect the frequency of the product within the population, which in turn would modify its price following a co-evolutionary dynamic.

@@

\xrc{close or remote in terms of what?}

@@

\xrc{I would move this discussion to the coevolution section? Besides, I would ass something such as "a large percentage of datasets used for this purpose are material objects found in archaeological contexts (i.e. pottery). Most of them are usually represented following a neutral model, but how this affects its weight inside the market system?"}

However, a limitation of the previous works is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the data analysed. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data which is then qualitatively compared to the real historical data.

\xrc{what data is being used?}

By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data. The Agent Based Models (ABM) methodology offers a way to produce this kind of model~\cite{lake_trends_2014}.

\xrc{I would add a paragraph briefly explaining what is an ABM}

@@

\xrc{The Pueblo model is an example, but there are tons of other ones...why is this chosen? After the paragraph explaining ABM I would say "As an example of its potential, a detailed model of the prehispanic blablabla..."}

\xrc{Didn't Brughmans have a model of roman trade?}

@@

\xrc{What if we move the text to the intro and remove this subsection?}

@@

\xrc{This first paragraph should briefly explain the trade component?}

@@

\xrc{I would use cultural transmission instead of imitation?}

@@

\xrc{how do agents trade?}

@@

\xrc{Not sure about the name...the sections below are describing the model so why are they in different section?}

@@

The first mechanism of the model aims at reproducing a scenario of unbiased transmission, where each product is a cultural trait without intrinsic positive or negative weight \cite{bentley_random_2004,bentley_specialisation_2005,mesoudi_random_2009}. 
Under this hypothesis, the process of \emph{production} is not relevant, and as a consequence, it does not modify the content of the quantity vector of the agents.

Unbiased transmission is implemented using ``random copy'': each agent has a low probability to pick randomly one agent among all and copy its vector of values. The \emph{innovation} process, termed ``unbounded'', is triggered with a low probability and draw a new random value to replace an element $v^i_j$. The imitation and innovation probabilities are presented with other parameters in table~\ref{tab:parameters}.

The neutral hypothesis states that the ``random copy'' transmission and the ``unbounded'' innovation process used under a fixed population size leads to a \emph{power law} distribution of the frequency for the different cultural variants. This distribution is characterized by a small number of very frequent traits and a large number of very rare traits. Since this hypothesis has been verified in previous works implementing similar \emph{imitation} and \emph{innovation} mechanisms, we expect in this work the observation of a similar ``power law' distribution.

\xrc{I would remove the previous sentence: it is obvious that it works right?}

More formally, this distribution is formalised as : $$P(v)=C/v^\alpha $$, where $v$ is the number of time a variant has been repeated, $P(v)$ the probability to find that variant, $C$ a constant, $\alpha$ a variable.

@@

\xrc{What do you mean by convergence? Equilibrium?}

@@

\xrc{score is fitness? Not sure if we should use fitness here, as we are already using an evolutionary framework. Also, we could use payoff.}
The \emph{score} $s^i$ of an agent $i$ reflects the ability of this agent to obtain the products it needs. It is maximum when the quantity vector of an agent is equal to its need vector and lower proportionally to the distance between the need vector and the quantity vector.  It is formally computed as follow for agent $i$:

@@

This function ensures that each good count for the same proportion \xrc{weight?} in the final score (i.e.: managing to get only the right amount of a good with a high ``need'' value will not give a better score to the agent). The values of $s^i$ obtained for different need values are depicted in figure~\ref{fig:fit}.

@@

\xrc{so no random copy? It should be clear that agents copy values based on random copy, and copy strategy based on payoff biased transmission (if i'm not mistaken).}

@@

On top of producing goods, the agents have the ability to exchange them. During the trading phase \xrc{we should explain the phases of the model before mentioning the trading phase} the value associated to a good by an agent corresponds to the subjective price of the good for this agent. Briefly summarised, for each good that it does not produce, an agent will trade with the first parter that offer an acceptable trade, i.e. the agent that proposes a satisfiable ratio between the other product and the product produced by the agent.

@@

\xrc{Add some 'scheduling' definition where this is shown. Also it should be justified (market is faster than cultural change?)}

@@

\todo{upload and put an url, complete cite mare nostrum, put nb of hours per run (or the total)}
All experiments are performed with the Pandora open source framework ~\cite{rubiocampillo_2014}. We have make the source code of our project available online~\footnote{put ulr}. \xrc{I always put this in the acknowledgements to avoid the technical details while describing the experiments}

All simulations presented are performed in the MareNostrumIII super computer~\cite{}, and took XXXh per run.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{ll|lllll}
		Agents Number & & Our result& & Bentley et al 2004\\
			&$\mu$ & $\alpha$ & SD&$\alpha$&SD\\\hline

		N = 250 
			&0.004&1.53&0.03&1.54&0.02\\
			&0.008&1.55&0.03&1.55&0.01\\
			&0.016&1.57&0.02&1.57&0.01\\
			&0.064&1.66&0.01&1.67&0.01\\\hline
		N= 500
		&0.004&1.50&0.02&1.53&0.03\\
		&0.008&1.52&0.02&1.57&0.01\\
		&0.016&1.55&0.03&1.61&0.04\\
		&0.064&1.78&0.08&1.81&0.10\\
	\end{tabular}
	\caption{Mean \& SD are calculate on 100 run.}
	\label{tab:mualpha}
\end{table}

@@

\xrc{ok but this is the experiment design right??}

@@

\xrc{these boxplots are confusing because they are small and contain a huge amount of information. What if we do a scatterplot with a given alpha, and on top of this a loess regression? It always works better to me when you have this kind of data; also, I would provide a figure for each type of score.}

@@

\xrc{discussion should go before conclusion. Also, this is not a disucussion but some comments on future stuff that could be done: what is the model explaining in terms of what was mentioned in the intro?}
%The questions we would like to ultimately study are for example: how does the hierarchical structure changes the adoption of a new product ? How delays and interruptions in the trade network change the culture of a society ?

@@

\section{Acknowledgements}

Funding for this work was provided by the ERC Advanced Grant EPNet (\todo{look for code}) and the SimulPast Consolider Ingenio project (CSD2010-00034) of the former Ministry for Science and Innovation of the Spanish Government.

The model was created using Pandora \cite{rubiocampillo_2014}. R was used for figures and statistical analysis \cite{rdev_2012}. The source code of the model is licensed under a GNU General Public License and will be available for download at publication.
